---
title: "COVID-19 Lineages"
description: Washington State DOH COVID-19 lineage extraction process
format: 
  html: 
    number-sections: false
    title-block-banner: false
    mainfont: Manrope
    toc: false
execute: 
  eval: false
---

This repo contains scripts that will pull lineages of interest from the CDC and transform them.

The repo will output a `.csv` file called `lineages.csv` that will be updated whenever new lineages are updated from the CDC. You can refresh/fetch this git repo for new changes and the `.csv` file will be updated.

To run the process:

1.  Run lineages script:

```{r filename="R console"}
source("scripts/lineages_public_repo.R")
```

2.  Run the lineages classification script:

```{r filename="R console"}
source("scripts/lineages_classification_for_test_repo.R")
```

## Variables

The variables produced by the scripts are used in the Sequencing and Variants Report. Here's a list:

| Variables               | Description                                                                                                                                    |
|---------------|---------------------------------------------------------|
| cdc_class               | Variable indicating VOC (variant of concern) or VBM (variant being monitored)                                                                  |
| who_name                | Variable indicating the WHO name                                                                                                               |
| doh_variant_name        | Grouping variable in WA DOH Sequencing and Variant report                                                                                      |
| hex_code                | Hex color for doh_variant_name group                                                                                                           |
| lineage_reporting_group | Variable indicating reporting group of lineage coded as: 1: Currently monitoring 2: Formerly monitoring 3: Formerly circulating, not monitored |
| report_table_name       | Variable name in numerical/pango form for table outputs                                                                                        |